Synthesising Recursive Functions for\\ First-Order Model Counting

### 2023-05-25 doc/talks/3_long/talk.tex
One more slide
Make the talk slightly longer

@@

% 15 min (8 slides, currently have 7) introduction
% 25 min (13 slides, currently have 10) presentation
% 5 min (3 slides, currently have 1) summary
% make sure there's a clear and coherent storyline

\author{Paulius Dilkas \\ \vspace{1cm} \tiny{Joint work with Vaishak Belle (Univeristy of Edinburgh, UK)}}
\title{Synthesising Recursive Functions for\\ First-Order Model Counting}
\institute{National University of Singapore, Singapore}
\date{PLSE Seminar, 21st April 2023}

@@

\begin{frame}[fragile]{(Some of the) Many Ways to Count}
  \begin{center}
    \begin{tikzcd}[remember picture]
      |[alias=WMI,visible on=<7->]|\WMI{} & |[visible on=<7->]|\WFOMI{} \\
      |[alias=WMC]|\#\SAT{}/\WMC{} \arrow[u,visible on=<7->] \ar[d,visible on=<8->] \ar[r,visible on=<5->] & |[visible on=<5->]|\textup{(W)\alert<9>{FOMC}} \arrow[u,visible on=<7->] \ar[d,visible on=<8->] \\
      |[alias=PBP,visible on=<8->]|\AMC{}, \pbp{} & |[alias=SP,visible on=<8->]|\SPPP{}
    \end{tikzcd}
  \end{center}
  \begin{tikzpicture}[overlay,remember picture]
    \begin{scope}[transform canvas={xshift=-3em}]
      \draw[->,draw=gray] ($(WMC)+(0,0.1)$)--($(WMI)+(0,0.5)$) node[midway,left,visible on=<7->] {\textcolor{gray}{$\sum \to \int$}};
      \draw[->,draw=gray] ($(WMC)-(0,0.1)$)--($(PBP)-(0,0.5)$) node[midway,left,visible on=<8->] {\textcolor{gray}{Generalising weights}};
    \end{scope}
    \begin{scope}[transform canvas={yshift=-2em}]
      \draw[->,draw=gray] ($(PBP)-(1,0)$)--($(SP)+(1,0)$) node[midway,below,visible on=<5->] {\textcolor{gray}{From propositional to first-order logic}};
    \end{scope}
  \vfill
    \begin{overprint}
      \onslide<1,2>
      \begin{block}{\#SAT~\textcolor{gray}{\parencite{DBLP:journals/tcs/Valiant79}}}
        \begin{itemize}
          \item Input formula: $\xx \lor \yy$
          \item Interpretations: $\emptyset$, $\{\, \xx \,\}$, $\{\, \yy \,\}$,
                $\{\, \xx, \yy \,\}$
          \item Models: $\{\, \xx \,\}$, $\{\, \yy \,\}$, $\{\, \xx, \yy \,\}$
          \item<2-> Answer (model count): 3
        \end{itemize}
      \end{block}
      \onslide<3,4>
      \begin{block}{Weighted Model Counting~\textcolor{gray}{\parencite{DBLP:journals/ai/ChaviraD08}}}
        \begin{itemize}
          \item Input formula: $\xx \lor \yy$
          \item Input weights: $\begin{aligned}[t]
                                  &w(\xx) = 0.3\text{, }w(\neg \xx) = 0.7,\\
                                  &w(\yy) = 0.2\text{, }w(\neg \yy) = 0.8
                                \end{aligned}$
          \item<4-> Answer (weighted model count):
                $w(\xx)w(\yy) + w(\xx)w(\neg \yy) + w(\neg \xx)w(\yy) = 0.44$
        \end{itemize}
      \end{block}
      \onslide<5,6>
      \begin{block}{(Weighted) (Symmetric) First-Order Model Counting\\
          \textcolor{gray}{\parencite{DBLP:conf/ijcai/BroeckTMDR11}}}
        \begin{itemize}
          \item Input formula: $\forall \xx \in \Delta\text{. }\pP(\xx)$
          \item Input weights: $w^{+}(\pP) = 0.3$, $w^{-}(\pP) = 0.7$
          \item Input domain size(s): $|\dDelta| = 2$
          \item<6-> Answer: ${(w^{+}(\pP))}^{|\dDelta|} = 0.09$
        \end{itemize}
      \end{block}
      \onslide<7>
      \begin{block}{Extensions to Continuous Domains}
        \begin{itemize}
          \item Weighted model integration
          \begin{itemize}
            \item \textcolor{gray}{\parencite{DBLP:conf/ijcai/BellePB15}}
          \end{itemize}
          \item Weighted first-order model integration
          \begin{itemize}
            \item \textcolor{gray}{\parencite{DBLP:conf/uai/FeldsteinB21}}
          \end{itemize}
        \end{itemize}
      \end{block}
      \onslide<8>
      \begin{block}{Generalisations of the Weight Function}
        \begin{itemize}
          \item Algebraic model counting
          \begin{itemize}
            \item \textcolor{gray}{\parencite{DBLP:journals/japll/KimmigBR17}}
            \item From $\mathbb{R}_{\ge 0}$ to commutative semirings
          \end{itemize}
          \item Pseudo-Boolean projection~\textcolor{gray}{\parencite{DBLP:conf/sat/DilkasB21}}
          \begin{itemize}
            \item Weights not necessarily on literals
          \end{itemize}
          \item Semiring programming~\textcolor{gray}{\parencite{DBLP:journals/ijar/BelleR20}}
        \end{itemize}
      \end{block}
    \end{overprint}
% maybe mention #CSP and other non-SAT ways of counting (SMT?, counting paths in
% automata and directed graphs)

@@

\begin{frame}{Exact Algorithms for FOMC}
  \begin{itemize}
    \item \alert{ForcLift}~\textcolor{gray}{\parencite{DBLP:conf/ijcai/BroeckTMDR11}}
    \begin{itemize}
      \item knowledge compilation to \alert{FO d-DNNF}
    \end{itemize}
    \item \alert{L2C}~\textcolor{gray}{\parencite{DBLP:conf/kr/KazemiP16}}
    \begin{itemize}
      \item knowledge compilation to \alert{C++} code
    \end{itemize}
    \item \alert{Alchemy}~\textcolor{gray}{\parencite{DBLP:journals/cacm/GogateD16}}
    \begin{itemize}
      \item \alert{DPLL}-style search
    \end{itemize}
    \item \alert{FastWFOMC}~\textcolor{gray}{\parencite{DBLP:conf/uai/BremenK21}}
    \begin{itemize}
      \item knowledge compilation to \alert{sd-DNNF}
    \end{itemize}
  \end{itemize}
  \pause
  \vfill
  \begin{block}{Our Contribution}
    \centering
    \begin{tikzpicture}[align=center,node distance=0.1cm]
      \node[label=below:{ForcLift}] (forclift) {\includegraphics[width=0.2\paperwidth]{forklift.png}};
      \node[right=of forclift] (plus) {\Huge$+$};
      \node[right=of plus,label=below:{Recursion}] (recursion) {\includegraphics[width=0.2\paperwidth]{matryoshka.jpg}};
      \node[right=of recursion] (equal) {\Huge$=$};
      \node[right=of equal,label=below:{Crane}] (crane) {\includegraphics[width=0.2\paperwidth]{crane.jpg}};
    \end{tikzpicture}
  \end{block}
\end{frame}

@@

% an example on partial injection (i.e., bipartite matching) counting
\begin{frame}{How These Rules Fit Together (1/5)}
  \begin{tikzpicture}[overlay,remember picture]
    \node[draw,rounded rectangle,fill=color2!50,anchor=center,above=0.5cm] at (current page.center) (rules) {Generalised domain recursion};
    \node[above=of rules] (before) {$\begin{gathered}
        \forall x \in \alert{\Gamma}\text{. }\forall y, z \in \Delta\text{. }y \ne z \Rightarrow \neg\pP(x, y) \lor \neg\pP(x, z)\\
        \forall w,x \in \alert{\Gamma}\text{. }\forall y \in \Delta\text{. }w \ne x \Rightarrow \neg\pP(w, y) \lor \neg\pP(x, y)
      \end{gathered}$};
    \node[below=of rules] (after) {$\begin{gathered}
        \forall y, z \in \Delta\text{. }y \ne z \Rightarrow \neg\pP(\alert{c}, y) \lor \neg\pP(\alert{c}, z)\\
        \forall x \in \Gamma\text{. }\forall y,z \in \Delta\text{. }y \ne z \land \alert{x \ne c} \Rightarrow \neg\pP(x, y) \lor \neg\pP(x, z)\\
        \forall x \in \Gamma\text{. }\forall y \in \Delta\text{. }\alert{x \ne c} \Rightarrow \neg\pP(\alert{c}, y) \lor \neg\pP(x, y)\\
        \forall w \in \Gamma\text{. }\forall y \in \Delta\text{. }\alert{w \ne c} \Rightarrow \neg\pP(w, y) \lor \neg\pP(\alert{c}, y)\\
        \forall w,x \in \Gamma\text{. }\forall y \in \Delta\text{. }w \ne x \land \alert{w \ne c} \land \alert{x \ne c} \Rightarrow \neg\pP(w, y) \lor \neg\pP(x, y)
      \end{gathered}$};
    \begin{scope}[on background layer]
      \node[anchor=north,fill=greenish,minimum width=11.1cm,minimum height=0.7cm] at (before.north) {};
      \node[anchor=south,fill=blueish,minimum width=11.1cm,minimum height=0.7cm] at (before.south) {};
      \node[anchor=north,fill=greenish,minimum width=11.1cm,minimum height=1.3cm] at (after.north) {};
      \node[anchor=south,fill=blueish,minimum width=11.1cm,minimum height=2cm] at (after.south) {};
    \end{scope}
    \draw[-Latex,color=color2,ultra thick,shorten <=0.1cm,shorten >=0.1cm] (before) -- (rules);
    \draw[-Latex,color=color2,ultra thick,shorten <=0.1cm,shorten >=0.1cm] (rules) -- (after);
  \end{tikzpicture}
\end{frame}

@@

\begin{frame}{Resulting Improvements to Counting Functions}
  Let \alert{$\Gamma$} and \alert{$\Delta$} be two sets with cardinalities
  \alert{$|\Gamma| = m$} and \alert{$|\Delta| = n$}.

  Our new rules enable Crane to efficiently count \alert{$\Gamma \to \Delta$}
  functions such as:
  \begin{itemize}
    \item injections in \alert{$\Theta(mn)$} time
    \begin{itemize}
      \item by hand: \alert{$\Theta(m)$}
    \end{itemize}
    \item partial injections in \alert{$\Theta(mn)$} time
    \begin{itemize}
      \item by hand: \alert{$\Theta({\min\{\, m, n \,\}}^2)$}
    \end{itemize}
    \item bijections in \alert{$\Theta(m)$} time
    \begin{itemize}
      \item \alert{optimal!}%\pause
      %\item In comparison, FastWFOMC scales as \alert{$\Omega(m^{4})$}.
    \end{itemize}
  \end{itemize}
  % \begin{tikzpicture}[remember picture,overlay]
  %   \node[draw,rectangle,color=gray] at ([xshift=-3cm]current page.east) {\includegraphics{plot}};
  % \end{tikzpicture}
\end{frame}

>>>>

\newcommand\Wider[2][3em]{%
\makebox[\linewidth][c]{%
  \begin{minipage}{\dimexpr\textwidth+#1\relax}
  \raggedright#2
  \end{minipage}%
  }%
}

@@

\author{Paulius Dilkas \\ \vspace{1cm} \tiny{Joint work with Vaishak Belle (Univeristy of Edinburgh, UK)}}
\title{Synthesising Recursive Functions for\\ First-Order Model Counting}
\institute{National University of Singapore, Singapore}
\date{DTAI Seminar, 26th May 2023}

@@

\begin{frame}[fragile]{(Some of the) Many Ways to Count}
  \begin{center}
    \begin{tikzcd}[remember picture]
      |[alias=WMI,visible on=<7->]|\WMI{} & |[visible on=<7->]|\WFOMI{} \\
      |[alias=WMC]|\#\SAT{}/\WMC{} \arrow[u,visible on=<7->] \ar[d,visible on=<8->] \ar[r,visible on=<5->] & |[visible on=<5->]|\textup{(W)\alert<9>{FOMC}} \arrow[u,visible on=<7->] \ar[d,visible on=<8->] \\
      |[alias=PBP,visible on=<8->]|\AMC{}, \pbp{} & |[alias=SP,visible on=<8->]|\SPPP{}
    \end{tikzcd}
  \end{center}
  \begin{tikzpicture}[overlay,remember picture]
    \begin{scope}[transform canvas={xshift=-3em}]
      \draw[->,draw=gray] ($(WMC)+(0,0.1)$)--($(WMI)+(0,0.5)$) node[midway,left,visible on=<7->] {\textcolor{gray}{$\sum \to \int$}};
      \draw[->,draw=gray] ($(WMC)-(0,0.1)$)--($(PBP)-(0,0.5)$) node[midway,left,visible on=<8->] {\textcolor{gray}{Generalising weights}};
    \end{scope}
    \begin{scope}[transform canvas={yshift=-2em}]
      \draw[->,draw=gray] ($(PBP)-(1,0)$)--($(SP)+(1,0)$) node[midway,below,visible on=<5->] {\textcolor{gray}{From propositional to first-order logic}};
    \end{scope}
  \end{tikzpicture}
  \vfill
    \begin{overprint}
      \onslide<1,2>
      \begin{block}{\#SAT~\textcolor{gray}{\parencite{DBLP:journals/tcs/Valiant79}}}
        \begin{itemize}
          \item Input formula: $\xx \lor \yy$
          \item Interpretations: $\emptyset$, $\{\, \xx \,\}$, $\{\, \yy \,\}$,
                $\{\, \xx, \yy \,\}$
          \item Models: $\{\, \xx \,\}$, $\{\, \yy \,\}$, $\{\, \xx, \yy \,\}$
          \item<2-> Answer (model count): 3
        \end{itemize}
      \end{block}
      \onslide<3,4>
      \begin{block}{Weighted Model Counting~\textcolor{gray}{\parencite{DBLP:journals/ai/ChaviraD08}}}
        \begin{itemize}
          \item Input formula: $\xx \lor \yy$
          \item Input weights: $\begin{aligned}[t]
                                  &w(\xx) = 0.3\text{, }w(\neg \xx) = 0.7,\\
                                  &w(\yy) = 0.2\text{, }w(\neg \yy) = 0.8
                                \end{aligned}$
          \item<4-> Answer (weighted model count):
                $w(\xx)w(\yy) + w(\xx)w(\neg \yy) + w(\neg \xx)w(\yy) = 0.44$
        \end{itemize}
      \end{block}
      \onslide<5,6>
      \begin{block}{(Weighted) (Symmetric) First-Order Model Counting\\
          \textcolor{gray}{\parencite{DBLP:conf/ijcai/BroeckTMDR11}}}
        \begin{itemize}
          \item Input formula: $\forall \xx \in \Delta\text{. }\pP(\xx)$
          \item Input weights: $w^{+}(\pP) = 0.3$, $w^{-}(\pP) = 0.7$
          \item Input domain size(s): $|\dDelta| = 2$
          \item<6-> Answer: ${(w^{+}(\pP))}^{|\dDelta|} = 0.09$
        \end{itemize}
      \end{block}
      \onslide<7>
      \begin{block}{Extensions to Continuous Domains}
        \begin{itemize}
          \item Weighted model integration
          \begin{itemize}
            \item \textcolor{gray}{\parencite{DBLP:conf/ijcai/BellePB15}}
          \end{itemize}
          \item Weighted first-order model integration
          \begin{itemize}
            \item \textcolor{gray}{\parencite{DBLP:conf/uai/FeldsteinB21}}
          \end{itemize}
        \end{itemize}
      \end{block}
      \onslide<8>
      \begin{block}{Generalisations of the Weight Function}
        \begin{itemize}
          \item Algebraic model counting
          \begin{itemize}
            \item \textcolor{gray}{\parencite{DBLP:journals/japll/KimmigBR17}}
            \item From $\mathbb{R}_{\ge 0}$ to commutative semirings
          \end{itemize}
          \item Pseudo-Boolean projection~\textcolor{gray}{\parencite{DBLP:conf/sat/DilkasB21}}
          \begin{itemize}
            \item Weights not necessarily on literals
          \end{itemize}
          \item Semiring programming~\textcolor{gray}{\parencite{DBLP:journals/ijar/BelleR20}}
        \end{itemize}
      \end{block}
    \end{overprint}
\end{frame}

@@

\begin{frame}{More Formally: What Is the Input?}
  \[
    \forall x \in \alert<2>{\Gamma}\text{. }\forall y, z \in \alert<2>{\Delta}\text{. }\pP(x, y) \land \pP(x, z) \Rightarrow \alert<4>{y=z}
  \]
  \begin{block}{\alert<2>{Many-Sorted} \alert<3>{Function-Free} First-Order Logic with \alert<4>{Equality}}
  \begin{itemize}
    \item<5-> Any number of variables
    \item<5-> All variables are bound
    \item<6-> \alert{$\forall$} and \alert{$\exists$} quantifiers can be nested arbitrarily deep
    \item<6-> All domains are finite
    \item<6-> Predicates can have any arity
  \end{itemize}
  \end{block}
\end{frame}

\begin{frame}{Exact Algorithms for FOMC}
  \begin{itemize}
    \item \alert{ForcLift}~\textcolor{gray}{\parencite{DBLP:conf/ijcai/BroeckTMDR11}}
    \begin{itemize}
      \item knowledge compilation to \alert{FO d-DNNF}
    \end{itemize}
    \item \alert{L2C}~\textcolor{gray}{\parencite{DBLP:conf/kr/KazemiP16}}
    \begin{itemize}
      \item knowledge compilation to \alert{C++} code
    \end{itemize}
    \item \alert{Alchemy}~\textcolor{gray}{\parencite{DBLP:journals/cacm/GogateD16}}
    \begin{itemize}
      \item \alert{DPLL}-style search
    \end{itemize}
    \item \alert{FastWFOMC}~\textcolor{gray}{\parencite{DBLP:conf/uai/BremenK21}}
    \begin{itemize}
      \item knowledge compilation to \alert{sd-DNNF}
    \end{itemize}
  \end{itemize}
  \pause
  \vfill
  \begin{block}{Our Contribution}
    \centering
    \begin{tikzpicture}[align=center,node distance=0.1cm]
      \node[label=below:{ForcLift}] (forclift) {\includegraphics[width=0.2\paperwidth]{forklift.png}};
      \node[right=of forclift] (plus) {\Huge$+$};
      \node[right=of plus,label=below:{Recursion}] (recursion) {\includegraphics[width=0.2\paperwidth]{matryoshka.jpg}};
      \node[right=of recursion] (equal) {\Huge$=$};
      \node[right=of equal,label=below:{Crane}] (crane) {\includegraphics[width=0.2\paperwidth]{crane.jpg}};
    \end{tikzpicture}
  \end{block}
\end{frame}
% NOTE: explain what knowledge compilation is

@@

\begin{frame}{How These Rules Fit Together (1/5)}
  \begin{tikzpicture}[overlay,remember picture]
    \node[draw,rounded rectangle,fill=color2!50,anchor=center,above=0.5cm] at (current page.center) (rules) {Generalised domain recursion};
    \node[above=of rules] (before) {$\begin{gathered}
        \forall x \in \alert{\Gamma}\text{. }\forall y, z \in \Delta\text{. }y \ne z \Rightarrow \neg\pP(x, y) \lor \neg\pP(x, z)\\
        \forall w,x \in \alert{\Gamma}\text{. }\forall y \in \Delta\text{. }w \ne x \Rightarrow \neg\pP(w, y) \lor \neg\pP(x, y)
      \end{gathered}$};
    \node[below=of rules] (after) {$\begin{gathered}
        \forall y, z \in \Delta\text{. }y \ne z \Rightarrow \neg\pP(\alert{c}, y) \lor \neg\pP(\alert{c}, z)\\
        \forall x \in \Gamma\text{. }\forall y,z \in \Delta\text{. }y \ne z \land \alert{x \ne c} \Rightarrow \neg\pP(x, y) \lor \neg\pP(x, z)\\
        \forall x \in \Gamma\text{. }\forall y \in \Delta\text{. }\alert{x \ne c} \Rightarrow \neg\pP(\alert{c}, y) \lor \neg\pP(x, y)\\
        \forall w \in \Gamma\text{. }\forall y \in \Delta\text{. }\alert{w \ne c} \Rightarrow \neg\pP(w, y) \lor \neg\pP(\alert{c}, y)\\
        \forall w,x \in \Gamma\text{. }\forall y \in \Delta\text{. }w \ne x \land \alert{w \ne c} \land \alert{x \ne c} \Rightarrow \neg\pP(w, y) \lor \neg\pP(x, y)
      \end{gathered}$};
    \begin{scope}[on background layer]
      \node[anchor=north,fill=greenish,minimum width=11.1cm,minimum height=0.7cm] at (before.north) {};
      \node[anchor=south,fill=blueish,minimum width=11.1cm,minimum height=0.7cm] at (before.south) {};
      \node[anchor=north,fill=greenish,minimum width=11.1cm,minimum height=1.3cm] at (after.north) {};
      \node[anchor=south,fill=blueish,minimum width=11.1cm,minimum height=2cm] at (after.south) {};
    \end{scope}
    \draw[-Latex,color=color2,ultra thick,shorten <=0.1cm,shorten >=0.1cm] (before) -- (rules);
    \draw[-Latex,color=color2,ultra thick,shorten <=0.1cm,shorten >=0.1cm] (rules) -- (after);
  \end{tikzpicture}
\end{frame}

@@

\begin{frame}{Resulting Improvements to Counting Functions}
  \Wider[3em]{ Let \alert{$\Gamma$} and \alert{$\Delta$} be two sets with
    cardinalities \alert{$|\Gamma| = m$} and \alert{$|\Delta| = n$}.

    Our new rules enable Crane to efficiently count \alert{$\Gamma \to \Delta$}
    functions such as: \vspace{1em}
    \begin{itemize}
            \setlength\itemsep{1em}
      \item injections in \alert{$\Theta(mn)$} time
            \begin{itemize}
              \item by hand: \alert{$\Theta(m)$}
            \end{itemize}
      \item partial injections in \alert{$\Theta(mn)$} time
            \begin{itemize}
              \item by hand: \alert{$\Theta({\min\{\, m, n \,\}}^2)$}
            \end{itemize}
      \item bijections in \alert{$\Theta(m)$} time
            \begin{itemize}
              \item \alert{optimal!}\pause
              \item In comparison, FastWFOMC scales as \alert{$\Omega(m^{4})$}.
            \end{itemize}
    \end{itemize}
  }
  \begin{tikzpicture}[remember picture,overlay]
    \node[draw,rectangle,color=gray] at ([xshift=-3.3cm,yshift=-0.2cm]current page.east) {\includegraphics{plot}};
  \end{tikzpicture}
\end{frame}
